Rising from the Ashes II: The Bar-driven Abundance Bimodality of the Milky Way

### 2024-10-19 paper/ms.tex
Updates to paper

\begin{abstract}
    The Milky Way hosts at least two modes in its present day distribution of Fe and $\alpha$-elements. The exact cause of this bimodality is disputed, but one class of explanations involves the merger between the Milky Way and a relatively massive satellite (Gaia-Sausage-Enceladus) at $z\sim2$. However, reproducing this bimodality in simulations is not straightforward, with conflicting results on the prevalance, morphology, and mechanism behind multimodality. We present a case study of a galaxy in the Illustris TNG50 simulation which undergoes a sequence of starburst, brief quiescence, and then rejuvenation. This sequence results in a pronounced abundance bimodality after a post-processing adjustment of the \alphaFe{} of old stars which is designed to mimic a higher star formation efficiency in dense gas. The high- and low-$\alpha$ sequences are neatly separated in time by the brief quiescent period. The quiescent period in this galaxy is not associated with a merger but by the formation of a bar followed by AGN activity. This galaxy indicates a novel scenario in which the $\alpha$-bimodality in the Milky Way is caused by the formation of the bar via AGN-induced quenching. We predict that abundance bimodalities should be more common in barred as opposed to unbarred galaxies.
  \end{abstract}

@@

\section{Introduction}\label{sec:intro}
The stellar surface abundances of most elements retain the composition of their natal gas cloud. Therefore, the present-day distribution of their abundances encodes the chemical history of a galaxy's gas phase. Two classes of elements have received particular attention in the Milky Way due to their disparate formation channels: iron-peak (such as Fe) and $\alpha$-elements (produced through the $\alpha$-process, such as O and Mg). Enrichment of iron-peak elements is primarily through Type~Ia and Type~II supernovae (SNe), whereas enrichment of $\alpha$-elements is primarily only through Type~II SNe. In the Milky Way, there is a well-established bimodality with separate high- and low-$\alpha$ sequences shown in Figure~\ref{fig:fig1} \citep{1996ASPC...92..307G,1998A&A...338..161F,2004AN....325....3F,2006MNRAS.367.1329R,2011A&A...535L..11A,2012A&A...545A..32A,2014A&A...562A..71B,2014ApJ...796...38N,2020MNRAS.493.2952H}.

@@

In \citet{2024arXiv240707985B} (hereafter Paper~I), we proposed an alternate scenario for the formation of the bimodality, driven by a brief ($\sim300\,\Myr$) quiescent period in the Galaxy's history (an observational claim with a broader period of quiescence was made in \citet{2016A&A...589A..66H}). This period has two effects. First, the reduced SFR lowers the number of $\alpha$-producing Type~II SNe relative to Type~Ia\footnote{These SNe occur on different timescales (10s of Myr after star formation for Type~II, as compared to 100s of Myr to Gyrs for Type~Ia), so if the SFR drops, the number of Type~II SNe drops faster than Type~Ia.}, causing a sharp decline in the \alphaFe{} ratio in the star-forming gas. Second, the lower SFR results in fewer stars forming in the region between the high- and low-$\alpha$ sequences, reducing the occurrence of this transitional population in present-day observations. The combination of these two effects creates the valley between the high- and low-$\alpha$ sequences. This mechanism resembles two-phase infall models that incorporate a temporary halt in star formation \citep[][and references therein]{2024arXiv240511025S}, though the quiescent period in Paper~I is much shorter.

In Paper~I, we used idealized simulations of a galaxy merger that triggered the quiescent period. However, in that work we argued the merger was unnecessary. In this work, we study a subhalo, or galaxy, from the Illustris TNG50 simulation demonstrating this. This galaxy exhibits the sequence of events presented in Paper~I after a simple post-processing step which increases the \alphaFe{} of old star particles. This post-processing is motivated by recent work demonstrating that the star formation efficiency (SFE) of the dense gas present at high-$z$ is too low in the TNG model \citep{2024arXiv240909121H}. The galaxy undergoes a brief quiescent period which neatly separates a high- and low-$\alpha$ sequence. However, instead of being preceded by a merger, the quenching is preceded by apparent bar-induced AGN activity. Therefore, this work serves as a verification that the scenario in Paper~I is possible in cosmological simulations and does not require a merger.

@@

We then further considered a dataset of stellar ages from the APOKASC-3 catalog \citep{2024arXiv241000102P}. This catalog uses a combination of APOGEE spectroscopic parameters and \textit{Kepler} time series photometry to compute astroseismic ages. Using only stars with $25\%$ age uncertainties (taken as the maximum of the upper and lower uncertainty), we cross-match this catalog to our larger sample from ASPCAP which results in a sample of 1201 stars.

@@

The colored vertical regions are indicated at $\FeH=-0.75$, $-0.5$, $-0.25$, and $0\,\dex$ in the Milky Way, and at bins $0.25\,\dex$ higher in the simulations. The lower panels show 1D histograms of \MgFe{} in bins centered on these values. The bins have width $0.2$/$0.05\,\dex$, which is reflected in the width of the colored, shaded regions in the upper panels. The rationale for the higher plotted \FeH{} in the simulations reflects the empirical location of the bimodalities. The Milky Way shows a clear bimodal population, with a high-$\alpha$ sequence most clearly distinct from the low-$\alpha$ sequence at low metallicity. The two sequences merge around solar metallicity.

@@

In the lower row, the left panel shows data from the Milky Way, while the middle and right panels correspond to the fiducial and $\alpha$-enhanced galaxy, respectively. For the simulations, we introduce Gaussian errors of $12.5\%$ in age and $0.01,\dex$ in \MgFe{}. These error values align with the observational uncertainties from the APOKASC-3 and APOGEE datasets, which the Milky Way panel (lower left) is based on (see Appendix~\ref{app:obs_err} for detailed plots of observational errors). While the error in \MgFe{} is insignificant, the age error ($1.5\,\Gyr$ at $10\,\Gyr$) blurs the distribution, particularly around the dashed line that marks the transition between sequences. Despite this, the $\alpha$-enhanced galaxy still shows two distinct populations, although their ages now overlap more significantly.

Interestingly, the Milky Way distribution (lower left panel) exhibits similarities to the $\alpha$-enhanced galaxy. Specifically, in the $\FeH=-0.5$ and $-0.25$ bins (blue and orange, respectively), there appear to be two populations that overlap in age but remain distinct in \MgFe{}. These are the bins where the bimodality is most pronounced (as seen in the upper left panel of Figure~\ref{fig:fig1}).

@@

The lower panel illustrates the growth of the galaxy's virial mass ($M_{200}$). Early on ($t<4\,\Gyr$), $M_{200}$ increases roughly linearly, reaching about $2 \times 10^{12}\,\Msun$. After this, the mass remains relatively stable until jumps occur around $t\sim10$ and $\sim12\,\Gyr$, indicative of mergers. These late-time mergers raise the virial mass to $5 \times 10^{12},\Msun$, well above the typical Milky Way estimate of $1$--$1.5\times 10^{12},\Msun$ \citep[e.g.][]{2016ARA&A..54..529B}. However, during the high- to low-$\alpha$ transition, the virial mass remains consistent with a Milky Way progenitor, making this galaxy a suitable analog. There are no mergers related to the earlier quiescent period around $t\sim3.2\,\Gyr$, as no major mass jumps are observed during this time. The lower right panel shows the period between $t=2$ and $5\,\Gyr$.

\subsection{Sequence of Events}\label{ssec:sequence_of_events}
\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig/alyssa.pdf}
  \caption{Surface density projections of gas (top row) and star particles (middle row) in our galaxy across snapshots at different times during the high- to low-$\alpha$ transition. Below the projections is a plot showing the SFR, BH accretion rate (in units of $\dot{M}_{\textrm{BH}}/\dot{M}_{\textrm{edd}}$), and bar strength ($A_2/A_0$ for $R<2$ kpc). Time ranges from $\sim2.4\,\textrm{Gyr}$ to $\sim3.6\,\textrm{Gyr}$, corresponding to redshifts from $z\sim2.7$ to $z\sim1.8$. A sequence of events in which the bar strengthens, BH accretion increases, and SFR declines is seen, and is described more fully in the text.}
  \label{fig:seq}
\end{figure*}

@@

This Figure shows the following sequence of events:
\begin{enumerate}
    \item Bar forms: A steady increase in the bar strength, as indicated by $A_2/A_0$ for star particles with $R<2\,\kpc$, from $\sim0.05$ to $0.4$ starting around $2.8\,\textrm{Gyr}$. This rise is accompanied by the appearance of elongated features in the gas and stars consistent with a bar.
    \item BH accretion increases: Following the increase in bar strength by about a snapshot ($\sim150\,\Myr$ here), the BH accretion rate ($\dot{M}_{\textrm{BH}}/\dot{M}_{\textrm{edd}}$) shows a significant spike, rising from a minimum of $\sim0.08$ at $t=2.84\,\Gyr$ to a maximum of $\sim0.29$ at $t=3.13\,\Gyr$ for one snapshot.
    \item SFR declines: The SFR declines starting from a maximum of $54.3\,\Msunyr$ at $t=2.84\,\Gyr$ down to $3.6\,\Msunyr$ at $t=3.28\,\Gyr$. In the next snapshot at $t=3.45\,\Gyr$ the SFR recovers to $12.2\,\Msunyr$. Figure~\ref{fig:history} shows that it reaches its second maximum of $30.9\,\Msunyr$ at $4.5\,\Gyr$.
\end{enumerate}

@@

Once the $\alpha$-enhancement post-processing has been done, the galaxy that we have studied in this work is consistent with this scenario. The vertical dashed line in Figures~\ref{fig:alpha} and \ref{fig:history} denotes the transition between the high- and low-$\alpha$ sequences. We place it at a lookback time of $10.6\,\Gyr$ or a cosmic time of $\sim3.2\,\Gyr$ or $z\sim2.05$.\footnote{The transition between the high- and low-$\alpha$ sequences occurs approximately $1\,\Gyr$ before $z=1.5$, where the $\alpha$-enhancement begins. This, combined with the fact that not all of the subhalos in our sample display bimodalities (Appendix~\ref{app:rand_fig1}), shows that the $\alpha$-enhancement is not the direct cause of the bimodality.} The upper right panel of Figure~\ref{fig:alpha} shows 2D histograms of \MgFe{} vs. star particle age in fixed bins of \FeH{} (each color is a different \FeH{} bin, and an offset is given to \MgFe{} to prevent overlap). Stars formed before the dashed line had a steep decline of \MgFe{} with time, while the relation in star particles formed after the dashed line is much more flat. At the $\alpha$-transition, we see a gap in the $\FeH=-0.5$ and $-0.25$ bins.

The age gap in each \FeH{} bin of Figure~\ref{fig:alpha} is contemperaneous with a minimum in the global SFR. We can see in the upper panel of Figure~\ref{fig:history} that this dashed line lies almost exactly at the point of a local minimum in the SFH. This minimum, which is $\sim3\,\Msunyr$, is 10--15$\times$ smaller than the maxima before and after it.

During this period of suppressed SF, we argue that the lack of enrichment of Type~II relative to Type~Ia SNe leads to a lower rate of Mg production. Therefore, \MgFe{} rapidly declines. This, combined with a lack of SF in the first place, leads to a scarcity of star particles in the intermediate region between the high- and low-$\alpha$ sequences. A more in-depth explanation is given in Section~4.1 in Paper~I.

@@

It is also worth noting that in both the fiducial and $\alpha$-enhanced galaxy show a slight rebound effect in \MgFe{}. The star particles which form directly after the dashed line have a slightly lower \MgFe{} than stars which form after. This was also predicted in Figure~9 of Paper~I, where it was argued that during the period of suppressed star formation, the \alphaFe{} ratio of the star-forming gas drops sharply due to the reduced contribution of Type~II relative to Type~Ia SNe. Later, the \alphaFe{} of the gas will recover when the SFR also recovers, but there is a brief window when old, low-$\alpha$ stars can form. A similar behavior was seen in the one-zone models with bursty SFHs in \citet{2020MNRAS.498.1364J}.

@@

We demonstrate the impact of the SFE on the \alphaFe{} ratio using a simple one-zone chemical evolution model with the publicly available code \texttt{VICE}. The details of our setup is given in Section~\ref{ssec:onezone_met}. We vary the SFE of the model ($\dot{\Sigma}_{\textrm{star}}/\Sigma_{\textrm{gas}}$), and examine the impact on the \MgFe{} ratio as a function of time. We find that a higher SFE does lead to a more rapid reduction in \MgFe{}. The rate of decrease in \MgFe{}, at its maximum, varies between $\sim-0.25$, $-0.35$, and $-0.5\,\dex/\Gyr$ in the $\textrm{SFE}=0.2$, $1$, and $2\,\Gyr^{-1}$ models, respectively. For our post-processing, we assumed an additional decrease rate of $0.1\,\dex/\Gyr$. Such a difference is well within the range of \MgFe{} slopes seen in our different $\tau_{\textrm{dep}}$ models, implying a factor of only $\sim2$ to $5$ in the SFE is needed to reach our post-processing slope.

@@

This work adds further support to a scenario in which a quiescent period in the Milky Way's past is a plausible explanation for the Milky Way's abundance bimodality. We argued in Paper~I that the GSE merger could trigger this period. In this work we have argued that the formation of the Milky Way's bar could be responsible. In external galaxies, we predict that the presence of a bar is correlated with the presence of an $\alpha$-bimodality.

>>>>

\begin{abstract}
    The Milky Way hosts at least two modes in its present day distribution of Fe and $\alpha$-elements. The exact cause of this bimodality is disputed, but one class of explanations involves the merger between the Milky Way and a relatively massive satellite (Gaia-Sausage-Enceladus) at $z\sim2$. However, reproducing this bimodality in simulations is not straightforward, with conflicting results on the prevalance, morphology, and mechanism behind multimodality. We present a case study of a galaxy in the Illustris TNG50 simulation which undergoes subsequent phases of starburst, brief quiescence, and then rejuvenation. This scenario results in a pronounced abundance bimodality after a post-processing adjustment of the \alphaFe{} of old stars designed to mimic a higher star formation efficiency in dense gas. The high- and low-$\alpha$ sequences are neatly separated in time by the brief quiescent period, which is not associated with a merger but by the formation of a bar followed by AGN activity. This galaxy indicates a novel scenario in which the $\alpha$-bimodality in the Milky Way is caused by the formation of the bar via AGN-induced quenching. In addition to a stellar age gap in the Milky Way, we predict that abundance bimodalities should be more common in barred as opposed to unbarred galaxies.
  \end{abstract}

@@

\section{Introduction}\label{sec:intro}
The stellar surface abundances of most elements retain the composition of their natal gas cloud. Therefore, the present-day distribution of their abundances encodes the chemical history of a galaxy's gas phase. Two classes of elements have received particular attention in the Milky Way due to their disparate formation channels: iron-peak (such as Fe) and $\alpha$-elements (produced through the $\alpha$-process, such as O and Mg). Enrichment of iron-peak elements is primarily through Type~Ia and Type~II supernovae (SNe), whereas enrichment of $\alpha$-elements is primarily only through Type~II SNe. In the Milky Way, there is a well-established bimodality with separate high- and low-$\alpha$ sequences as shown in the upper left panel of Figure~\ref{fig:fig1} \citep{1996ASPC...92..307G,1998A&A...338..161F,2004AN....325....3F,2006MNRAS.367.1329R,2011A&A...535L..11A,2012A&A...545A..32A,2014A&A...562A..71B,2014ApJ...796...38N,2020MNRAS.493.2952H}.

@@

In \citet{2024arXiv240707985B} (hereafter Paper~I), we proposed an alternate scenario for the formation of the bimodality, driven by a brief ($\sim300\,\Myr$) quiescent period in the Galaxy's history (an observational claim with a longer period of quiescence was made in \citet{2016A&A...589A..66H}). This period has two effects. First, the reduced SFR lowers the number of $\alpha$-producing Type~II SNe relative to Type~Ia\footnote{These SNe occur on different timescales (10s of Myr after star formation for Type~II, as compared to 100s of Myr to Gyrs for Type~Ia), so if the SFR drops, the number of Type~II SNe drops faster than Type~Ia.}, causing a sharp decline in the \alphaFe{} ratio in the star-forming gas. Second, the lower SFR results in fewer stars forming in the intermediate region between the high- and low-$\alpha$ sequences, reducing the occurrence of this transitional population in present-day observations. The combination of these two effects creates the valley between the high- and low-$\alpha$ sequences. This mechanism resembles two-phase infall models that incorporate a temporary halt in star formation \citep[][and references therein]{2024arXiv240511025S}, though the quiescent period in Paper~I is much shorter.

In Paper~I, we used idealized simulations of a galaxy merger that triggered the quiescent period. However, in that work we argued that it was the quiescent period, not the merger, which was necessary to produce a bimodality. In this work, we study a subhalo, or galaxy, from the Illustris TNG50 simulation demonstrating this. This galaxy exhibits the sequence of events presented in Paper~I after a post-processing step, motivated by recent work showing that the star formation efficiency (SFE) of dense gas at high-$z$ is too low in the TNG model \citep{2024arXiv240909121H}, which increases the \alphaFe{} of old star particles. The galaxy undergoes a brief quiescent period which neatly separates a high- and low-$\alpha$ sequence. However, instead of being preceded by a merger, the quenching is preceded by apparent bar-induced AGN activity. Therefore, this work serves as a verification that the scenario in Paper~I is possible in cosmological simulations and does not require a merger.

@@

We then further considered a dataset of stellar ages from the APOKASC-3 catalog \citep{2024arXiv241000102P}. This catalog uses a combination of APOGEE spectroscopic parameters and \textit{Kepler} time series photometry to compute astroseismic ages. Using only stars with $25\%$ age uncertainties (taken as the maximum of the upper and lower uncertainty), we cross-match this catalog to our larger sample from ASPCAP which results in a sample of 2525 stars.

@@

The colored vertical regions are indicated at $\FeH=-0.75$, $-0.5$, $-0.25$, and $0\,\dex$ in the Milky Way, and at bins $0.25\,\dex$ higher in the simulations. The lower panels show 1D histograms of \MgFe{} in bins centered on these values. The bins have width $0.1\,\dex$, which is reflected in the width of the colored, shaded regions in the upper panels. The rationale for the higher plotted \FeH{} in the simulations reflects the empirical location of the bimodalities. The Milky Way shows a clear bimodal population, with a high-$\alpha$ sequence most clearly distinct from the low-$\alpha$ sequence at low metallicity. The two sequences merge around solar metallicity.

@@

In the lower row, the left panel shows data from the Milky Way, while the middle and right panels correspond to the fiducial and $\alpha$-enhanced galaxy, respectively. For the simulations, we introduce Gaussian errors of $12.5\%$ in age and $0.01\,dex$ in \MgFe{}. These error values align with the observational uncertainties from the APOKASC-3 and APOGEE datasets, which the Milky Way panel (lower left) is based on (see Appendix~\ref{app:obs_err} for detailed plots of observational errors). While the error in \MgFe{} is insignificant, the age error ($1.25\,\Gyr$ at $10\,\Gyr$) blurs the distribution, particularly around the dashed line that marks the transition between sequences. Despite this, the $\alpha$-enhanced galaxy still shows two distinct populations, although their ages now overlap more significantly.

The Milky Way distribution (lower left panel) exhibits some similarities to the $\alpha$-enhanced galaxy. Specifically, in the $\FeH=-0.5$ and $-0.25$ bins (blue and orange, respectively), there appear to be two populations that overlap in age but remain distinct in \MgFe{}. These are the bins where the bimodality is most pronounced (as seen in the upper left panel of Figure~\ref{fig:fig1}). In the Milky Way there also appears to be a population of $\alpha$-rich stars down to low ages (see discussion in Section~\ref{ssec:compare_obs}).

@@

The lower panel illustrates the growth of the galaxy's virial mass ($M_{200}$). Early on ($t<4\,\Gyr$), $M_{200}$ increases roughly linearly, reaching about $2 \times 10^{12}\,\Msun$. After this, the mass remains relatively stable until jumps occur around $t\sim10$ and $\sim12\,\Gyr$, indicative of mergers. These late-time mergers raise the virial mass to $5 \times 10^{12},\Msun$, well above the typical Milky Way estimate of $1$--$1.5\times 10^{12},\Msun$ \citep[e.g.][]{2016ARA&A..54..529B}. However, during the high- to low-$\alpha$ transition, the virial mass remains consistent with a Milky Way progenitor, making this galaxy a suitable analog. There are no mergers related to the earlier quiescent period around $t\sim3.2\,\Gyr$, as no major mass jumps are observed during this time. The lower right panel shows the lack of mergers in more detail during the period between $t=2$ and $5\,\Gyr$.

\subsection{Sequence of Events}\label{ssec:sequence_of_events}
\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig/alyssa.pdf}
  \caption{\textbf{Quiescence separating the high- and low-$\alpha$ sequences is preceded by AGN activity associated with bar formation.} Surface density projections of gas (top row) and star particles (middle row) in our galaxy across snapshots at different times during the high- to low-$\alpha$ transition. Below the projections is a plot showing the SFR, BH accretion rate (in units of $\dot{M}_{\textrm{BH}}/\dot{M}_{\textrm{edd}}$), and bar strength ($A_2/A_0$ for $R<2$ kpc). Time ranges from $\sim2.4\,\textrm{Gyr}$ to $\sim3.6\,\textrm{Gyr}$, corresponding to redshifts from $z\sim2.7$ to $z\sim1.8$. A sequence of events in which the bar strengthens, BH accretion increases, and SFR declines is seen, and is described more fully in the text.}
  \label{fig:seq}
\end{figure*}

@@

This figure shows the following sequence of events:
\begin{enumerate}
    \item Bar forms: A steady increase in the bar strength, as indicated by $A_2/A_0$ for star particles with $R<2\,\kpc$, from $\sim0.05$ to $0.4$ starting around $2.8\,\textrm{Gyr}$. This rise is accompanied by the appearance of elongated features in the gas and stars consistent with a bar.
    \item BH accretion increases: Following the increase in bar strength by about a snapshot ($\sim150\,\Myr$ here), the BH accretion rate ($\dot{M}_{\textrm{BH}}/\dot{M}_{\textrm{edd}}$) shows a significant spike, rising from a minimum of $\sim0.08$ at $t=2.84\,\Gyr$ to a maximum of $\sim0.29$ at $t=3.13\,\Gyr$ for one snapshot.
    \item SFR declines: The SFR declines starting from a maximum of $54.3\,\Msunyr$ at $t=2.84\,\Gyr$ down to $3.6\,\Msunyr$ at $t=3.28\,\Gyr$. In the next snapshot at $t=3.45\,\Gyr$ the SFR recovers to $12.2\,\Msunyr$. Figure~\ref{fig:history} shows that it reaches its second maximum of $30.9\,\Msunyr$ at $4.5\,\Gyr$.
\end{enumerate}

@@

Once the $\alpha$-enhancement post-processing has been done, the galaxy that we have studied in this work is consistent with the quiescent scenario proposed in Paper~I. The distribution of \MgFe{} with star particle age is a useful way to test this scenario, which we plot in the upper right panel of Figure~\ref{fig:alpha} in fixed bins of \FeH{} (each color is a different \FeH{} bin). The vertical dashed line in Figures~\ref{fig:alpha} and \ref{fig:history} marks the transition between the high- and low-$\alpha$ sequences. We place this transition at a lookback time of $10.6,\Gyr$, corresponding to a cosmic time of approximately $3.2,\Gyr$ or redshift $z\sim2.05$. \footnote{The transition between the high- and low-$\alpha$ sequences occurs approximately $1\,\Gyr$ before $z=1.5$, where the $\alpha$-enhancement begins. This, combined with the fact that not all of the subhalos in our sample display bimodalities (Appendix~\ref{app:rand_fig1}), shows that the $\alpha$-enhancement is not the direct cause of the bimodality.} Star particles formed before this transition show a steep decline in \MgFe{} over time, with a drop of approximately $0.3\,\dex$ in just $2\,\Gyr$. In contrast, the stars formed after the dashed line exhibit a much flatter trend, with no noticeable gradient in \MgFe{} for the next couple of Gyr.

At the transition, we can see a gap in the $\FeH=-0.5$, $-0.25$, and $0\,\dex$ bins (orange, green, and red, respectively). The gap has a width of about $300\,\Myr$ in the $-0.5$ and $-0.25\,\dex$ bins, while it is more extended at about $750\,\Myr$ in the $0\,\dex$ bin. The age gap in each \FeH{} bin of Figure~\ref{fig:alpha} is contemperaneous with a minimum in the global SFR. We can see in the upper panel of Figure~\ref{fig:history} that this dashed line lies almost exactly at the point of a local minimum in the SFH. This minimum, which is $\sim3\,\Msunyr$, is 10--15$\times$ smaller than the maxima before and after it.

During this period of suppressed SF, we argue that the lack of enrichment of Type~II relative to Type~Ia SNe leads to a lower rate of Mg production. The typical lifetime of a Type~II SN progenitor in this model is $\sim40\,\Myr$ \citep{2018MNRAS.473.4077P}, and so the hundreds of Myr of suppressed SF is short enough to restrict the production of $\alpha$-elements. Therefore, \MgFe{} rapidly declines. This, combined with a lack of SF in the first place, leads to a scarcity of star particles in the intermediate region between the high- and low-$\alpha$ sequences. A more in-depth explanation is given in Section~4.1 in Paper~I.

@@

Notably, both the fiducial and $\alpha$-enhanced galaxy show a slight rebound effect in \MgFe{}. The star particles which form directly after the dashed line when the SFR has just recovered have a slightly lower \MgFe{} than stars which form later on, by about $0.1$ to $0.2\,\dex$. This was also predicted in Figure~9 of Paper~I, where it was argued that during the period of suppressed star formation, the \alphaFe{} ratio of the star-forming gas drops sharply due to the reduced contribution of Type~II relative to Type~Ia SNe. Later, the \alphaFe{} of the gas will recover when the SFR also recovers, but there is a brief window when old, low-$\alpha$ stars can form. A similar behavior was seen in the one-zone models with bursty SFHs in \citet{2020MNRAS.498.1364J}.

@@

We demonstrate the impact of the SFE on the \alphaFe{} ratio using a simple one-zone chemical evolution model with the publicly available code \texttt{VICE}. The details of our setup is given in Section~\ref{ssec:onezone_met}. We vary the SFE of the model ($\textrm{SFR}/M_{\textrm{gas}}$), and examine the impact on \MgFe{} as a function of time. We find that a higher SFE does lead to a more rapid reduction in \MgFe{}. The rate of decrease in \MgFe{}, at its maximum, varies between $\sim-0.25$, $-0.35$, and $-0.5\,\dex/\Gyr$ in the $\textrm{SFE}=0.2$, $1$, and $2\,\Gyr^{-1}$ models, respectively. For our post-processing, we assumed an additional decrease rate of $0.1\,\dex/\Gyr$. Such a difference is well within the range of \MgFe{} slopes seen in our different $\tau_{\textrm{dep}}$ models, implying a factor of only $\sim2$ to $5$ in the SFE is needed to reach our post-processing slope.

@@

This work adds further support to a scenario in which a quiescent period in the Milky Way's past is a plausible explanation for the Milky Way's abundance bimodality. We argued in Paper~I that the GSE merger could trigger this period. In this work we have argued that the formation of the Milky Way's bar could be responsible. Regardless of this scenario's relevance to the Milky Way, we also predict that the presence of a bar in external galaxies is correlated with the presence of an $\alpha$-bimodality.
